%
\begin{isabellebody}%
\def\isabellecontext{Sessions}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isatagtheory
\isacommand{theory}\isamarkupfalse%
\ Sessions\isanewline
\isakeyword{imports}\ Base\isanewline
\isakeyword{begin}%
\endisatagtheory
{\isafoldtheory}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isamarkupchapter{Isabelle sessions and build management \label{ch:session}%
}
\isamarkuptrue%
%
\begin{isamarkuptext}%
An Isabelle \emph{session} consists of a collection of related
  theories that are associated with an optional formal document (see
  also \chref{ch:present}).  There is also a notion of persistent
  \emph{heap image} to capture the state of a session, similar to
  object-code in compiled programming languages.

  Thus the concept of session resembles that of a ``project'' in
  common IDE environments, but our specific name emphasizes the
  connection to interactive theorem proving: the session wraps-up the
  results of user-interaction with the prover in a persistent form.

  \medskip Application sessions are built on a given parent session.
  The resulting hierarchy eventually leads to some major object-logic
  session, notably \isa{{\isaliteral{22}{\isachardoublequote}}HOL{\isaliteral{22}{\isachardoublequote}}}, which itself is based on \isa{{\isaliteral{22}{\isachardoublequote}}Pure{\isaliteral{22}{\isachardoublequote}}} as the common root.

  Processing sessions may take considerable time.  The tools for
  Isabelle build management help to organize this efficiently,
  including support for parallel build jobs --- in addition to the
  multithreaded theory and proof checking that is already provided by
  the prover process.%
\end{isamarkuptext}%
\isamarkuptrue%
%
\isamarkupsection{Session ROOT specifications \label{sec:session-root}%
}
\isamarkuptrue%
%
\begin{isamarkuptext}%
Session specifications reside in files called \verb|ROOT|
  within certain directories, such as the home locations of registered
  Isabelle components or additional project directories given by the
  user.

  The ROOT file format follows the lexical conventions of the
  \emph{outer syntax} of Isabelle/Isar, see also
  \cite{isabelle-isar-ref}.  This defines common forms like
  identifiers, names, quoted strings, verbatim text, nested comments
  etc.  The grammar for a single \hyperlink{syntax.session-entry}{\mbox{\isa{session{\isaliteral{5F}{\isacharunderscore}}entry}}} is given as
  syntax diagram below; each ROOT file may contain multiple session
  specifications like this.

  Isabelle/jEdit (\secref{sec:tool-jedit}) includes a simple editing
  mode \verb|isabelle-root| for session ROOT files.

  \begin{railoutput}
\rail@begin{2}{\indexdef{}{syntax}{session\_entry}\hypertarget{syntax.session-entry}{\hyperlink{syntax.session-entry}{\mbox{\isa{session{\isaliteral{5F}{\isacharunderscore}}entry}}}}}
\rail@term{\isa{\isakeyword{session}}}[]
\rail@nont{\isa{spec}}[]
\rail@term{\isa{{\isaliteral{3D}{\isacharequal}}}}[]
\rail@bar
\rail@nextbar{1}
\rail@nont{\hyperlink{syntax.name}{\mbox{\isa{name}}}}[]
\rail@term{\isa{{\isaliteral{2B}{\isacharplus}}}}[]
\rail@endbar
\rail@nont{\isa{body}}[]
\rail@end
\rail@begin{2}{\isa{body}}
\rail@bar
\rail@nextbar{1}
\rail@nont{\isa{description}}[]
\rail@endbar
\rail@bar
\rail@nextbar{1}
\rail@nont{\isa{options}}[]
\rail@endbar
\rail@plus
\rail@nextplus{1}
\rail@cnont{\isa{theories}}[]
\rail@endplus
\rail@bar
\rail@nextbar{1}
\rail@nont{\isa{files}}[]
\rail@endbar
\rail@end
\rail@begin{2}{\isa{spec}}
\rail@nont{\hyperlink{syntax.name}{\mbox{\isa{name}}}}[]
\rail@bar
\rail@nextbar{1}
\rail@term{\isa{{\isaliteral{21}{\isacharbang}}}}[]
\rail@endbar
\rail@bar
\rail@nextbar{1}
\rail@nont{\isa{groups}}[]
\rail@endbar
\rail@bar
\rail@nextbar{1}
\rail@nont{\isa{dir}}[]
\rail@endbar
\rail@end
\rail@begin{2}{\isa{groups}}
\rail@term{\isa{{\isaliteral{28}{\isacharparenleft}}}}[]
\rail@plus
\rail@nont{\hyperlink{syntax.name}{\mbox{\isa{name}}}}[]
\rail@nextplus{1}
\rail@endplus
\rail@term{\isa{{\isaliteral{29}{\isacharparenright}}}}[]
\rail@end
\rail@begin{1}{\isa{dir}}
\rail@term{\isa{\isakeyword{in}}}[]
\rail@nont{\hyperlink{syntax.name}{\mbox{\isa{name}}}}[]
\rail@end
\rail@begin{1}{\isa{description}}
\rail@term{\isa{\isakeyword{description}}}[]
\rail@nont{\hyperlink{syntax.text}{\mbox{\isa{text}}}}[]
\rail@end
\rail@begin{1}{\isa{options}}
\rail@term{\isa{\isakeyword{options}}}[]
\rail@nont{\isa{opts}}[]
\rail@end
\rail@begin{3}{\isa{opts}}
\rail@term{\isa{{\isaliteral{5B}{\isacharbrackleft}}}}[]
\rail@plus
\rail@bar
\rail@nont{\hyperlink{syntax.name}{\mbox{\isa{name}}}}[]
\rail@term{\isa{{\isaliteral{3D}{\isacharequal}}}}[]
\rail@nont{\hyperlink{syntax.name}{\mbox{\isa{name}}}}[]
\rail@nextbar{1}
\rail@nont{\hyperlink{syntax.name}{\mbox{\isa{name}}}}[]
\rail@endbar
\rail@nextplus{2}
\rail@cterm{\isa{{\isaliteral{2C}{\isacharcomma}}}}[]
\rail@endplus
\rail@term{\isa{{\isaliteral{5D}{\isacharbrackright}}}}[]
\rail@end
\rail@begin{2}{\isa{theories}}
\rail@term{\isa{\isakeyword{theories}}}[]
\rail@bar
\rail@nextbar{1}
\rail@nont{\isa{opts}}[]
\rail@endbar
\rail@plus
\rail@nont{\hyperlink{syntax.name}{\mbox{\isa{name}}}}[]
\rail@nextplus{1}
\rail@endplus
\rail@end
\rail@begin{2}{\isa{files}}
\rail@term{\isa{\isakeyword{files}}}[]
\rail@plus
\rail@nont{\hyperlink{syntax.name}{\mbox{\isa{name}}}}[]
\rail@nextplus{1}
\rail@endplus
\rail@end
\end{railoutput}


  \begin{description}

  \item \isakeyword{session}~\isa{{\isaliteral{22}{\isachardoublequote}}A\ {\isaliteral{3D}{\isacharequal}}\ B\ {\isaliteral{2B}{\isacharplus}}\ body{\isaliteral{22}{\isachardoublequote}}} defines a new
  session \isa{{\isaliteral{22}{\isachardoublequote}}A{\isaliteral{22}{\isachardoublequote}}} based on parent session \isa{{\isaliteral{22}{\isachardoublequote}}B{\isaliteral{22}{\isachardoublequote}}}, with its
  content given in \isa{body} (theories and auxiliary source files).
  Note that a parent (like \isa{{\isaliteral{22}{\isachardoublequote}}HOL{\isaliteral{22}{\isachardoublequote}}}) is mandatory in practical
  applications: only Isabelle/Pure can bootstrap itself from nothing.

  All such session specifications together describe a hierarchy (tree)
  of sessions, with globally unique names.  By default, names are
  derived from parent ones by concatenation, i.e.\ \isa{{\isaliteral{22}{\isachardoublequote}}B{\isaliteral{5C3C646173683E}{\isasymdash}}A{\isaliteral{22}{\isachardoublequote}}}
  above.  Cumulatively, this leads to session paths of the form \isa{{\isaliteral{22}{\isachardoublequote}}X{\isaliteral{5C3C646173683E}{\isasymdash}}Y{\isaliteral{5C3C646173683E}{\isasymdash}}Z{\isaliteral{5C3C646173683E}{\isasymdash}}W{\isaliteral{22}{\isachardoublequote}}}.  Note that in the specification,
  \isa{B} is already such a fully-qualified name, while \isa{{\isaliteral{22}{\isachardoublequote}}A{\isaliteral{22}{\isachardoublequote}}}
  is the new base name.

  \item \isakeyword{session}~\isa{{\isaliteral{22}{\isachardoublequote}}A{\isaliteral{21}{\isacharbang}}\ {\isaliteral{3D}{\isacharequal}}\ B{\isaliteral{22}{\isachardoublequote}}} indicates a fresh start
  in the naming scheme: the session is called just \isa{{\isaliteral{22}{\isachardoublequote}}A{\isaliteral{22}{\isachardoublequote}}} instead
  of \isa{{\isaliteral{22}{\isachardoublequote}}B{\isaliteral{5C3C646173683E}{\isasymdash}}A{\isaliteral{22}{\isachardoublequote}}}.  Here the name \isa{{\isaliteral{22}{\isachardoublequote}}A{\isaliteral{22}{\isachardoublequote}}} should be
  sufficiently long to stand on its own in a potentially large
  library.

  \item \isakeyword{session}~\isa{{\isaliteral{22}{\isachardoublequote}}A\ {\isaliteral{28}{\isacharparenleft}}groups{\isaliteral{29}{\isacharparenright}}{\isaliteral{22}{\isachardoublequote}}} indicates a
  collection of groups where the new session is a member.  Group names
  are uninterpreted and merely follow certain conventions.  For
  example, the Isabelle distribution tags some important sessions by
  the group name \isa{{\isaliteral{22}{\isachardoublequote}}main{\isaliteral{22}{\isachardoublequote}}}.  Other projects may invent their own
  conventions, which requires some care to avoid clashes within this
  unchecked name space.

  \item \isakeyword{session}~\isa{{\isaliteral{22}{\isachardoublequote}}A{\isaliteral{22}{\isachardoublequote}}}~\isakeyword{in}~\isa{{\isaliteral{22}{\isachardoublequote}}dir{\isaliteral{22}{\isachardoublequote}}}
  specifies an explicit directory for this session.  By default,
  \isakeyword{session}~\isa{{\isaliteral{22}{\isachardoublequote}}A{\isaliteral{22}{\isachardoublequote}}} abbreviates
  \isakeyword{session}~\isa{{\isaliteral{22}{\isachardoublequote}}A{\isaliteral{22}{\isachardoublequote}}}~\isakeyword{in}~\isa{{\isaliteral{22}{\isachardoublequote}}A{\isaliteral{22}{\isachardoublequote}}}.  This
  accommodates the common scheme where some base directory contains
  several sessions in sub-directories of corresponding names.  Another
  common scheme is \isakeyword{session}~\isa{{\isaliteral{22}{\isachardoublequote}}A{\isaliteral{22}{\isachardoublequote}}}~\isakeyword{in}~\verb|"."| to refer to the current
  directory of the ROOT file.

  All theories and auxiliary source files are located relatively to
  the session directory.  The prover process is run within the same as
  its current working directory.

  \item \isakeyword{description}~\isa{{\isaliteral{22}{\isachardoublequote}}text{\isaliteral{22}{\isachardoublequote}}} is a free-form
  annotation for this session.

  \item \isakeyword{options}~\isa{{\isaliteral{22}{\isachardoublequote}}{\isaliteral{5B}{\isacharbrackleft}}x\ {\isaliteral{3D}{\isacharequal}}\ a{\isaliteral{2C}{\isacharcomma}}\ y\ {\isaliteral{3D}{\isacharequal}}\ b{\isaliteral{2C}{\isacharcomma}}\ z{\isaliteral{5D}{\isacharbrackright}}{\isaliteral{22}{\isachardoublequote}}} defines
  separate options that are used when processing this session, but
  \emph{without} propagation to child sessions; see also
  \secref{sec:system-options}.  Note that \isa{{\isaliteral{22}{\isachardoublequote}}z{\isaliteral{22}{\isachardoublequote}}} abbreviates
  \isa{{\isaliteral{22}{\isachardoublequote}}z\ {\isaliteral{3D}{\isacharequal}}\ true{\isaliteral{22}{\isachardoublequote}}} for Boolean options.

  \item \isakeyword{theories}~\isa{{\isaliteral{22}{\isachardoublequote}}options\ names{\isaliteral{22}{\isachardoublequote}}} specifies a
  block of theories that are processed within an environment that is
  augmented further by the given options, in addition to the global
  session options given before.  Any number of blocks of
  \isakeyword{theories} may be given.  Options are only active for
  each \isakeyword{theories} block separately.

  \item \isakeyword{files}~\isa{{\isaliteral{22}{\isachardoublequote}}files{\isaliteral{22}{\isachardoublequote}}} lists additional source
  files that are somehow involved in the processing of this session.
  This should cover anything outside the formal content of the theory
  sources, say some auxiliary {\TeX} files that are required for
  document processing.  In contrast, files that are specified in
  formal theory headers as \hyperlink{keyword.uses}{\mbox{\isa{\isakeyword{uses}}}} need not be declared
  again.

  \end{description}%
\end{isamarkuptext}%
\isamarkuptrue%
%
\isamarkupsubsubsection{Examples%
}
\isamarkuptrue%
%
\begin{isamarkuptext}%
See \verb|~~/src/HOL/ROOT| for a diversity of practically
  relevant situations.%
\end{isamarkuptext}%
\isamarkuptrue%
%
\isamarkupsection{System build options \label{sec:system-options}%
}
\isamarkuptrue%
%
\begin{isamarkuptext}%
See \verb|~~/etc/options| for the main defaults provided by
  the Isabelle distribution.  Isabelle/jEdit (\secref{sec:tool-jedit})
  includes a simple editing mode \verb|isabelle-options| for
  this file-format.%
\end{isamarkuptext}%
\isamarkuptrue%
%
\isamarkupsection{Invoking the build process \label{sec:tool-build}%
}
\isamarkuptrue%
%
\begin{isamarkuptext}%
The \indexdef{}{tool}{build}\hypertarget{tool.build}{\hyperlink{tool.build}{\mbox{\isa{\isatt{build}}}}} utility invokes the build process for
  Isabelle sessions.  It manages dependencies between sessions,
  related sources of theories and auxiliary files, and target heap
  images.  Accordingly, it runs instances of the prover process with
  optional document preparation.  Its command-line usage
  is:\footnote{Isabelle/Scala provides the same functionality via
  \texttt{isabelle.Build.build}.}
\begin{ttbox} Usage: isabelle build [OPTIONS] [SESSIONS ...]

  Options are:
    -a           select all sessions
    -b           build heap images
    -d DIR       include session directory with ROOT file
    -g NAME      select session group NAME
    -j INT       maximum number of parallel jobs (default 1)
    -n           no build -- test dependencies only
    -o OPTION    override session configuration OPTION
                 (via NAME=VAL or NAME)
    -s           system build mode: produce output in ISABELLE_HOME
    -v           verbose

  Build and manage Isabelle sessions, depending on implicit
  ISABELLE_BUILD_OPTIONS="..."

  ML_PLATFORM="..."
  ML_HOME="..."
  ML_SYSTEM="..."
  ML_OPTIONS="..."
\end{ttbox}

  \medskip Isabelle sessions are defined via session ROOT files as
  described in (\secref{sec:session-root}).  The totality of sessions
  is determined by collecting such specifications from all Isabelle
  component directories (\secref{sec:components}), augmented by more
  directories given via options \verb|-d|~\isa{{\isaliteral{22}{\isachardoublequote}}DIR{\isaliteral{22}{\isachardoublequote}}} on the
  command line.  Each such directory may contain a session
  \texttt{ROOT} file and an additional catalog file \verb|etc/sessions| with further sub-directories (list of lines).  Note
  that a single \texttt{ROOT} file usually defines many sessions;
  catalogs are only required for extra scalability and modularity
  of large libraries.

  \medskip The subset of sessions to be managed is selected via
  individual \isa{{\isaliteral{22}{\isachardoublequote}}SESSIONS{\isaliteral{22}{\isachardoublequote}}} given as command-line arguments, or
  session groups that are specified via one or more options \verb|-g|~\isa{{\isaliteral{22}{\isachardoublequote}}NAME{\isaliteral{22}{\isachardoublequote}}}.  Option \verb|-a| selects all sessions.
  The build tool takes the hierarchy of sessions into account: the
  selected set of sessions is completed by including all ancestors
  according to the dependency structure.

  \medskip The build process depends on additional options that are
  passed to the prover session eventually, see also
  (\secref{sec:system-options}).  The settings variable \indexref{}{setting}{ISABELLE\_BUILD\_OPTIONS}\hyperlink{setting.ISABELLE-BUILD-OPTIONS}{\mbox{\isa{\isatt{ISABELLE{\isaliteral{5F}{\isacharunderscore}}BUILD{\isaliteral{5F}{\isacharunderscore}}OPTIONS}}}} allows to provide additional defaults.
  Moreover, the environment of system build options may be augmented
  on the command line via \verb|-o|~\isa{{\isaliteral{22}{\isachardoublequote}}NAME{\isaliteral{3D}{\isacharequal}}VALUE{\isaliteral{22}{\isachardoublequote}}} or
  \verb|-o|~\isa{{\isaliteral{22}{\isachardoublequote}}NAME{\isaliteral{22}{\isachardoublequote}}}, which abbreviates \verb|-o|~\isa{{\isaliteral{22}{\isachardoublequote}}NAME{\isaliteral{3D}{\isacharequal}}true{\isaliteral{22}{\isachardoublequote}}} for Boolean options.  Multiple occurrences
  of \verb|-o| on the command-line are applied in the given
  order.

  \medskip Option \verb|-b| ensures that heap images are
  produced for all selected sessions.  By default, images are only
  saved for inner nodes of the hierarchy of sessions, as required for
  other sessions to continue later on.

  \medskip Option \verb|-j| specifies the maximum number of
  parallel build jobs (prover processes).  Note that each process is
  subject to a separate limit of parallel threads, cf.\ system option
  \indexref{}{system option}{threads}\hyperlink{system option.threads}{\mbox{\isa{\isatt{threads}}}}.

  \medskip Option \verb|-s| enables \emph{system mode}, which
  means that resulting heap images and log files are stored in
  \verb|$ISABELLE_HOME/heaps| instead of the default location
  \hyperlink{setting.ISABELLE-OUTPUT}{\mbox{\isa{\isatt{ISABELLE{\isaliteral{5F}{\isacharunderscore}}OUTPUT}}}} (which is normally in \hyperlink{setting.ISABELLE-HOME-USER}{\mbox{\isa{\isatt{ISABELLE{\isaliteral{5F}{\isacharunderscore}}HOME{\isaliteral{5F}{\isacharunderscore}}USER}}}}, i.e.\ the user's home directory).

  \medskip Option \verb|-n| omits the actual build process
  after performing all dependency checks.  The return code indicates
  the status of the set of selected sessions.

  \medskip Option \verb|-v| enables verbose mode.%
\end{isamarkuptext}%
\isamarkuptrue%
%
\isamarkupsubsubsection{Examples%
}
\isamarkuptrue%
%
\begin{isamarkuptext}%
Build a specific logic image:
\begin{ttbox}
isabelle build -b HOLCF
\end{ttbox}

  Build the main group of logic images (as determined by the session
  ROOT specifications of the Isabelle distribution):
\begin{ttbox}
isabelle build -b -g main
\end{ttbox}

  Provide a general overview of the status of all Isabelle sessions,
  without building anything:
\begin{ttbox}
isabelle build -a -n -v
\end{ttbox}

  Build all sessions with HTML browser info and PDF document
  preparation:
\begin{ttbox}
isabelle build -a -o browser_info -o document=pdf
\end{ttbox}

  Build all sessions with a maximum of 8 prover processes and 4
  threads each (on a machine with many cores):

\begin{ttbox}
isabelle build -a -j8 -o threads=4
\end{ttbox}%
\end{isamarkuptext}%
\isamarkuptrue%
%
\isadelimtheory
%
\endisadelimtheory
%
\isatagtheory
\isacommand{end}\isamarkupfalse%
%
\endisatagtheory
{\isafoldtheory}%
%
\isadelimtheory
%
\endisadelimtheory
\isanewline
\end{isabellebody}%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "root"
%%% End:
