\chapter{More about Types}
\label{ch:more-types}

So far we have learned about a few basic types (for example \isa{bool} and
\isa{nat}), type abbreviations (\isacommand{types}) and recursive datatypes
(\isacommand{datatype}). This chapter will introduce more
advanced material:
\begin{itemize}
\item Pairs ({\S}\ref{sec:products}) and records ({\S}\ref{sec:records}),
and how to reason about them.
\item Type classes: how to specify and reason about axiomatic collections of
  types ({\S}\ref{sec:axclass}). This section leads on to a discussion of
  Isabelle's numeric types ({\S}\ref{sec:numbers}).  
\item Introducing your own types: how to define types that
  cannot be constructed with any of the basic methods
  ({\S}\ref{sec:adv-typedef}).
\end{itemize}

The material in this section goes beyond the needs of most novices.
Serious users should at least skim the sections as far as type classes.
That material is fairly advanced; read the beginning to understand what it
is about, but consult the rest only when necessary.

\index{pairs and tuples|(}
\input{document/Pairs}    %%%Section "Pairs and Tuples"
\index{pairs and tuples|)}

\input{document/Records}  %%%Section "Records"


\section{Type Classes} %%%Section
\label{sec:axclass}
\index{axiomatic type classes|(}
\index{*axclass|(}

The programming language Haskell has popularized the notion of type
classes: a type class is a set of types with a
common interface: all types in that class must provide the functions
in the interface.  Isabelle offers a similar type class concept: in
addition, properties (\emph{class axioms}) can be specified which any
instance of this type class must obey.  Thus we can talk about a type
$\tau$ being in a class $C$, which is written $\tau :: C$.  This is the case
if $\tau$ satisfies the axioms of $C$. Furthermore, type classes can be
organized in a hierarchy.  Thus there is the notion of a class $D$
being a subclass\index{subclasses} of a class $C$, written $D
< C$. This is the case if all axioms of $C$ are also provable in $D$.

In this section we introduce the most important concepts behind type
classes by means of a running example from algebra.  This should give
you an intuition how to use type classes and to understand
specifications involving type classes.  Type classes are covered more
deeply in a separate tutorial \cite{isabelle-classes}.

\subsection{Overloading}
\label{sec:overloading}
\index{overloading|(}

\input{document/Overloading}

\index{overloading|)}

\input{document/Axioms}

\index{type classes|)}
\index{*class|)}

\input{Types/numerics}             %%%Section "Numbers"

\input{document/Typedefs}    %%%Section "Introducing New Types"
